Demystifying the Computer Science \\ PhD Admission in the US\\{\large A Handbook for International \emph{and Domestic} Students

### 2024-08-20 demystify.tex
updates

International students, especially those from less well-known universities and smaller countries, often wonder if they should apply to a CS PhD program in the US. Specifically, they are worried about the \textbf{difficulty of getting admitted} and the \textbf{cost of graduate study} in the US. This section addresses these concerns.

@@

Second, PhD students in CS \emph{do not} need to worry about funding, especially at good R1
universities in the US. If you are admitted, you will almost certainly
receive \emph{full funding} (\autoref{sec:funding}) to support your study, including tuition,
health insurance, and stipend (i.e., you get paid for your study). Moreover, depending on the university,
you may even receive additional benefits such as summer pay, laptops, (conference/workshop) traveling.

@@

After you submit your PhD application (e.g., in late Fall of year $X$ for the Fall admission of year $X+1$), it will first be ``screened''
for general requirements, e.g., did you submit your transcripts and standard scores? did your reference writers submit their letters? Usually, this screening process is done through a central university system, i.e., not by CS faculty. If you pass this, your application will be forwarded to the CS department for further evaluation. If you don't,  the system will tell you what is missing and what you need to do. So pay attention to your email and check your application status regularly.

@@

\paragraph{Adcom} After screening applications are reviewed by a \emph{PhD admission committee} (\textbf{adcom}) that consists of faculty members in CS\footnote{In some cases the committee can involve affiliated faculty from different disciplines.}. These adcom members have a wide range of expertise and background to ensure diverse perspectives in the evaluation process. The size and the review load of the adcom depend on the department size. At Mason, the PhD adcom typically has 15--20 faculty, and each committee member is assigned to review about 30 applications. Note that most large schools, including Mason, have separate adcoms for MS programs.

@@

  \item[Adcom] Admission Committee. The group of faculty members who review applications and make admission decisions.
  \item [Adcom members] Faculty members who are part of the admission committee. These are the people who review applications and make admission decisions.
  \item[Adcom chair] The faculty member who leads the admission committee. This person is often not involved in individual admission decisions but oversees the process (e.g., resolve disputes, ensure fairness).
  \item[Advisor/Supervisor]: A faculty member who guides and mentors a PhD student throughout their research. This person typically plays a crucial role in a student's academic journey. In the US, the term ``adviser'' is more commonly used than ``supervisor''.
  \item[Cohort] A group of students who start a program at the same time and take classes together. You should get to know your cohort as they will be your colleagues and friends during your PhD.
  \item  [Deferred Admission] An option allowing admitted students to postpone their start date, typically by one year.
  \item [Diversity Statement] A document sometimes required in applications where applicants describe their contributions to diversity and inclusion, or how their background adds to the diversity of the academic community. This document is often required for faculty positions, but some universities also require it for graduate admissions.
  \item[Domestic Student] A student who is typically a US citizen or permanent resident (``greencard holder'').
        But sometimes it can also refer to students who did their undergrad (or MS) at a US university.
  \item[Field or Area] A specific research area, e.g., Machine Learning, Computer Vision, Software Engineering, etc.
  \item [In-state vs. Out-of-state] In-state tuition is the tuition rate for students who are residents of the state where the university is located. Out-of-state tuition is the tuition rate for students who are not residents of that state. In-state tuition is typically much lower than out-of-state tuition. PhD students typically do not have to pay tuition as it is covered by their funding, but this might be important for MS students.
  \item [Lab] Research Lab. A group of researchers (profs. and students) working on a common research area, e.g., a Software Engineering Lab.  Typically a lab is led by a single professor, but sometimes it can be co-led by multiple professors. For example, my lab is DynaROARS, which focuses on my own research area (software verification), but I am also part of the more general SWE Lab at Mason with other faculty in Software Engineering.

  \item[LoR] Letter of Recommendation. A letter written by a professor or supervisor that assesses your qualifications and potential for graduate study (\autoref{sec:lor}).
  \item[MS] Master of Science. A graduate degree that typically requires 2 years of study.
  \item[Open House] An event where admitted students visit the university to meet faculty, students, and staff, and learn more about the program. This is a great opportunity to get a feel for the university and the department and decide if it is a good fit for you.

@@

  \item[RA] Research Assistantship. A form of funding where you work as a research assistant for a professor.
  \item[TA] Teaching Assistantship. A form of funding where you work as a teaching assistant for a professor.
  \item[TOEFL] Test of English as a Foreign Language. A standardized test to measure the English language ability of non-native speakers.
  \item[IELTS] International English Language Testing System. A standardized test to measure the English language ability of non-native speakers.
  \item[GRE] Graduate Record Examination. A standardized test that is an admissions requirement for many graduate schools in the US. However, it is not required for most CS PhD programs (\autoref{sec:standard-tests}).

  \item [PI]  Principal Investigator. A lead researcher on a grant or research project, e.g., a PI on an NSF grant. Prospective students often use this term to refer to a professor they are interested in working with.  While this term is used frequently in online forums among students, we do not use it this way in academia, e.g., ``who is your PI?'' might only make sense to fellow Reddit applicants, but might not be understood by others, ``who is your adviser?'' is what you want to use.
    \item[STEM] Science, Technology, Engineering, and Mathematics. A term used to group together these academic disciplines. CS is considered a STEM field.
  \item[URM] Underrepresented Minority. Refers to groups that are underrepresented in higher education, e.g., Black, Hispanic, and Native American students.
  \item[NSF] National Science Foundation. A US government agency that supports research and education in science and engineering. NSF is a major source of funding for RAs in CS research. There are other agencies that support RAs but they might require US citizenship or permanent residency.
        \item [R1 vs. R2 Universities] Research 1 (R1) universities are universities with the highest level of research activity. Research 2 (R2) universities have a high level of research activity but not as high as R1 universities.  R1 universities are typically more prestigious and have more funding and resources for research.  R2 universities are also good but might have fewer resources and funding.
  \item [Rolling Admission] A process where applications are reviewed as they are received, and decisions are made throughout the admission cycle.  This is in contrast to a process where all applications are reviewed after the deadline and decisions are made at the same time. Rolling admission is more common for MS programs and less common for PhD programs (e.g., at Mason, MS is rolling and PhD is not).
  \item [Top-tier conferences and journals] These are the most prestigious venues for publishing research in CS.  For example, in Software Engineering, top-tier conferences include ICSE, FSE, and ASE and top-tier journals include TSE and TOSEM.  Publishing in these venues is highly competitive and prestigious, and can significantly improve your chances of getting admitted to a good PhD program.

>>>>

International students, especially those from less well-known universities and smaller countries, often wonder if they should apply to a CS PhD program in the US. Specifically, they are worried about (i) the \textbf{difficulty of getting admitted} and (ii) the \textbf{cost of graduate study} in the US. This section addresses these concerns.

@@

Second, PhD students in CS \emph{do not} need to worry about funding, especially at good R1
universities in the US. If you are admitted, you will almost certainly
receive \emph{full funding} (\autoref{sec:funding}) to support your study, including tuition,
health insurance, and stipend (i.e., you get paid for your study). Moreover, depending on the university,
you may even receive additional benefits such as summer pay, laptops, and traveling to conferences and workshops.

@@

After you submit your PhD application (e.g., in late Fall of year $X$ for the Fall admission of year $X+1$), it will first be checked
for general requirements, e.g., did you submit your transcripts and standard scores? did your reference writers submit their letters? Usually, this screening process is done through a central university system, i.e., not by CS faculty.

If you pass this, your application will be forwarded to the CS department for further evaluation. If you don't,  the system will tell you what is missing and what you need to do. So pay attention to your email and check your application status regularly.

@@

\paragraph{Adcom} After screening applications are reviewed by a PhD admission committee or \emph{adcom} that consists of faculty members in CS\footnote{In some cases the committee can involve affiliated faculty from different disciplines.}. These adcom members have a wide range of expertise and background to ensure diverse perspectives in the evaluation process. The size and the review load of the adcom depend on the department size. At Mason, the PhD adcom typically has 15--20 faculty, and each committee member is assigned to review about 30 applications. Note that most large schools, including Mason, have separate adcoms for MS programs.

@@

  \item[Adcom] Admission Committee. The group of faculty members who review applications and make admission decisions.
  \item [Adcom members] Faculty members who are part of the admission committee. These are the people who review applications and make admission decisions.
  \item[Adcom chair] The faculty member who leads the admission committee. This person is often not involved in individual admission decisions but oversees the process (e.g., resolve disputes, ensure fairness).
  \item[Advisor/Supervisor]: A faculty member who guides and mentors a PhD student throughout their research. This person typically plays a crucial role in a student's academic journey. In the US, the term ``adviser'' is more commonly used than ``supervisor''.
  \item[Cohort] A group of students who start a program at the same time and take classes together. You should get to know your cohort as they will be your colleagues and friends during your PhD.
  \item  [Deferred Admission] An option allowing admitted students to postpone their start date, typically by one year.
  \item [Diversity Statement] A document sometimes required in applications where applicants describe their contributions to diversity and inclusion, or how their background adds to the diversity of the academic community. This document is often required for faculty positions, but some universities also require it for graduate admissions.
  \item[Domestic Student] A student who is typically a US citizen or permanent resident (``greencard holder'').
        But sometimes it can also refer to students who did their undergrad (or MS) at a US university.
  \item[Field or Area] A specific research area, e.g., Machine Learning, Computer Vision, Software Engineering, etc.
  \item [In-state vs. Out-of-state] In-state tuition is the tuition rate for students who are residents of the state where the university is located. Out-of-state tuition is the tuition rate for students who are not residents of that state. In-state tuition is typically much lower than out-of-state tuition. PhD students typically do not have to pay tuition as it is covered by their funding, but this might be important for MS students.

  \item [Lab] Research Lab. A group of researchers (profs. and students) working on a common research area, e.g., a Software Engineering Lab.  Typically a lab is led by a single professor, but sometimes it can be co-led by multiple professors. For example, my lab is DynaROARS, which focuses on my own research area (software verification), but I am also part of the more general SWE Lab at Mason with other faculty in Software Engineering.

  \item[LoR] Letter of Recommendation. A letter written by a professor or supervisor that assesses your qualifications and potential for graduate study (\autoref{sec:lor}).

  \item[Major] A student's primary field of study, e.g., Computer Science, Electrical Engineering, Mathematics, etc. Often used in the context of undergraduate studies (e.g., ``what was your major in undergrad?'').

  \item[MS] Master of Science. A graduate degree that typically requires 2 years of study.
  \item[Open House] An event where admitted students visit the university to meet faculty, students, and staff, and learn more about the program. This is a great opportunity to get a feel for the university and the department and decide if it is a good fit for you.

@@

  \item[RA and TA] Research Assistantship (RA) is form of funding where you work on a research project for a professor. Teaching Assistantship (TA) is a form of funding where you help a a professor with teaching (e.g., grading assignments).

  \item[TOEFL/IELTS] Test of English as a Foreign Language. A standardized test to measure the English language ability of non-native speakers.
  \item[GRE] Graduate Record Examination. A standardized test that is an admissions requirement for many graduate schools in the US. However, it is not required for most CS PhD programs (\autoref{sec:standard-tests}).

  \item [PI]  Principal Investigator. A lead researcher on a grant or research project, e.g., a PI on an NSF grant. Prospective students often use this term to refer to a professor they are interested in working with.  While this term is used frequently in online forums among students, we do not use it this way in academia, e.g., ``who is your PI?'' might only make sense to fellow Reddit applicants, but might not be understood by others, ``who is your adviser?'' is what you want to use.
    \item[STEM] Science, Technology, Engineering, and Mathematics. A term used to group together these academic disciplines. CS is considered a STEM field.
  \item[URM] Underrepresented Minority. Refers to groups that are underrepresented in higher education, e.g., Black, Hispanic, and Native American students.
  \item[NSF] National Science Foundation. A US government agency that supports research and education in science and engineering. NSF is a major source of funding for RAs in CS research. There are other agencies that support RAs but they might require US citizenship or permanent residency.
        \item [R1/R2 Universities] Research 1 (R1) universities are universities with the highest level of research activity. Research 2 (R2) universities have a high level of research activity but not as high as R1 universities.  R1 universities are typically more prestigious and have more funding and resources for research.  R2 universities are also good but might have fewer resources and funding.
  \item [Rolling Admission] A process where applications are reviewed as they are received, and decisions are made throughout the admission cycle.  This is in contrast to a process where all applications are reviewed after the deadline and decisions are made at the same time. Rolling admission is more common for MS programs and less common for PhD programs (e.g., at Mason, MS is rolling and PhD is not).
  \item [Top-tier conferences and journals] These are the most prestigious venues for publishing research in CS.  For example, in Software Engineering, top-tier conferences include ICSE, FSE, and ASE and top-tier journals include TSE and TOSEM.  Publishing in these venues is highly competitive and prestigious, and can significantly improve your chances of getting admitted to a good PhD program.
